---
title: "Bayesian analysis of the packaged example data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian analysis of the packaged example data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

cores <- parallel::detectCores(logical = FALSE)
options(mc.cores = if(is.na(cores)) 4L else min(4L, cores))
```

`TruncComp` provides an experimental Bayesian interface,
`trunc_comp_bayes()`, for the no-covariate two-group comparison. This vignette
applies that interface to the packaged example data set and shows how to move
from the atom-coded endpoint to posterior summaries for the atom-probability
contrast, the non-atom mean contrast, and the coded-endpoint contrast. The
packaged example uses the default `continuous_support = "real_line"` pathway.
The interface also supports
`continuous_support = "positive_real"` when all non-atom outcomes are strictly
positive, and experimental bounded models through
`continuous_support = "bounded_continuous"` or
`continuous_support = "bounded_score"`. The bounded models use Beta mixture
kernels by default and can instead use logit-normal mixture kernels with
`bounded_kernel = "logit_normal"`.

The example data set is intentionally small. The Stan fit below uses four
chains with 500 warmup iterations and 1,000 post-warmup iterations per chain.
Inspect the sampler diagnostics and increase the sampling effort when they are
unsatisfactory.

## Load the example data

```{r}
library(TruncComp)

data("trunc_comp_example", package = "TruncComp")
example_data <- transform(
  trunc_comp_example,
  A = as.integer(Y != 0)
)

head(example_data, 10)
```

The packaged example uses `0` as the atom. Thus `A = 0` marks the substantive
event for which the continuous outcome is undefined, and `A = 1` marks rows
outside the atom.

Before fitting the Bayesian model, it is helpful to inspect the two groups on
the three scales that matter most:

- the atom probability
- the mean among the non-atom outcomes
- the mean of the coded endpoint

```{r}
arm_summary <- do.call(
  rbind,
  lapply(split(example_data, example_data$R), function(d) {
    data.frame(
      n = nrow(d),
      atom_probability = mean(d$A == 0L),
      outside_atom_probability = mean(d$A == 1L),
      mean_non_atom = mean(d$Y[d$A == 1L]),
      mean_coded_endpoint = mean(d$Y),
      row.names = NULL
    )
  })
)

arm_summary$arm <- c("control", "treatment")
arm_summary <- arm_summary[, c("arm", "n", "atom_probability",
                               "outside_atom_probability", "mean_non_atom",
                               "mean_coded_endpoint")]
arm_summary
```

This quick table already shows why the two-part decomposition is useful: one
group can look better on the non-atom mean while also having a worse atom
probability.

## Fit the Bayesian two-part model

The Bayesian interface uses the same unadjusted formula structure as the
frequentist interface. Its model has the following structure:

- the atom is modeled separately in each arm
- the non-atom outcomes are modeled with arm-specific truncated stick-breaking
  mixtures chosen by `continuous_support`
- bounded continuous and bounded score supports can use either Beta kernels or
  logit-normal kernels
- covariate adjustment is unavailable

```{r fit-bayes, message = FALSE, warning = FALSE}
fit_bayes <- trunc_comp_bayes(
  Y ~ R,
  atom = 0,
  data = example_data,
  chains = 4,
  iter_warmup = 500,
  iter_sampling = 1000,
  mixture_components = 6,
  seed = 20260418,
  refresh = 0
)

fit_bayes
```

The printed fit shows the four main treatment-effect summaries:

- `delta_atom`: difference in atom probability
- `mu_delta`: difference in means among non-atom outcomes
- `alpha_delta`: odds ratio for being outside the atom
- `delta`: coded-endpoint contrast

The stored posterior summary table is often the most convenient compact view:

```{r}
fit_bayes$summary_table
```

These reported point estimates are posterior medians. The `posterior_prob`
column is the posterior probability that the corresponding contrast exceeds its
null value, which is `0` for `delta_atom`, `mu_delta`, and `delta`, and `1` for
`alpha_delta`.

The model also stores arm-specific posterior summaries:

```{r}
fit_bayes$arm_table
```

Here `rho_0` and `rho_1` are the arm-specific atom probabilities, `pi_0` and
`pi_1` are the arm-specific probabilities of being outside the atom,
and `mu_0_c` and `mu_1_c` are the means among the non-atom outcomes.

## Posterior outcome densities

The Bayesian fit object can also be visualized through the arm-specific
posterior outcome densities implied by the fitted mixture. The plotted curve in
each panel summarizes the non-atom part of the posterior predictive law, so it
is weighted by the posterior probability of being outside the atom. The
dark-gray ribbon shows a pointwise credible interval, and the solid spike with
an arrowhead records the atom together with its posterior mean mass.
For this vignette, we supply a compact x-grid directly so the figure focuses on
the region where the example data live while still showing the atom at `0`.

```{r, fig.width=11, fig.height=5.5, out.width="100%"}
non_atom_range <- range(example_data$Y[example_data$A == 1])
density_x <- seq(
  min(fit_bayes$atom, non_atom_range[1]) - 0.3,
  non_atom_range[2] + 0.8,
  length.out = 300
)

posterior_density_plot(fit_bayes, x = density_x)
```

## Posterior predictive checks

The fit object supports posterior predictive checks for both parts of the
Bayesian model. The atom check compares the fitted Bernoulli atom model with the
atom indicators, while the non-atom check compares replicated draws from the
fitted mixture with the non-atom outcomes. The package also reports
discrepancy-based posterior predictive p-values for these two checks. These are
posterior predictive model-checking summaries rather than frequentist
hypothesis-test p-values.

```{r}
posterior_predictive_pvalues(fit_bayes)
```

```{r, fig.width=11, fig.height=5.5, out.width="100%"}
ppc_plots <- posterior_predictive_check(fit_bayes, seed = 1)
print(ppc_plots$atom)
print(ppc_plots$continuous)
```

## Credible intervals and custom posterior summaries

You can extract equal-tail credible intervals for the main contrasts with
`confint()`:

```{r}
confint(fit_bayes, parameter = c("delta_atom", "mu_delta", "delta"))
```

The fit stores full posterior draws in `fit_bayes$draws`, so it is easy to
compute custom posterior summaries. For example, if smaller atom probability
is preferable, then a treatment benefit on the atom component corresponds to
`delta_atom < 0` rather than `delta_atom > 0`.

```{r}
posterior_probabilities <- c(
  `P(delta > 0)` = mean(fit_bayes$draws$delta > 0),
  `P(mu_delta > 0)` = mean(fit_bayes$draws$mu_delta > 0),
  `P(delta_atom < 0)` = mean(fit_bayes$draws$delta_atom < 0)
)

posterior_probabilities
```

The stored posterior draws support study-specific summaries that are not part
of the default output.

## Diagnostics and practical tuning

The fit object stores a compact diagnostic summary:

```{r}
data.frame(
  divergences = fit_bayes$diagnostics$divergences,
  max_rhat = fit_bayes$diagnostics$max_rhat,
  min_bulk_ess = fit_bayes$diagnostics$min_bulk_ess,
  min_tail_ess = fit_bayes$diagnostics$min_tail_ess,
  core_ok = fit_bayes$diagnostics$core_ok,
  truncation_ok = fit_bayes$diagnostics$truncation_ok,
  diagnostic_ok = fit_bayes$diagnostics$diagnostic_ok
)
```

The displayed `max_rhat`, `min_bulk_ess`, `min_tail_ess`, and `core_ok` values
summarize the four reported treatment contrasts. For truncated stick-breaking
mixtures, `truncation_ok` separately summarizes ordinary sampler diagnostics
for the concentration parameters and the final component weights. Neither
summary assesses approximation error from the finite mixture representation.

Because this vignette uses a small packaged example data set, these
diagnostics should be interpreted as an illustration of the workflow rather
than as definitive evidence that the model is adequate for every
application. Increase the sampling effort when the diagnostics are
unsatisfactory, then inspect `summary(fit_bayes)` and
`fit_bayes$diagnostics` again.

For data with the same structure:

1. code the atom explicitly
2. fit `trunc_comp_bayes(Y ~ R, atom = ...)`
3. inspect `summary_table`, `arm_table`, and diagnostics
4. compute any study-specific posterior probabilities from `fit$draws`

## Positive-support continuous outcomes

If the non-atom outcomes are constrained to the positive real line, set
`continuous_support = "positive_real"`. That switches the continuous component
from a Gaussian mixture to a Gamma mixture while keeping the same atom model
and the same posterior estimands.

```{r fit-bayes-positive, message = FALSE, warning = FALSE}
positive_example <- simulate_truncated_data(
  20,
  f0 = function(n) stats::rgamma(n, shape = 4, rate = 2.5),
  f1 = function(n) stats::rgamma(n, shape = 5, rate = 2.2),
  pi0 = 0.7,
  pi1 = 0.6,
  atom = 0
)

fit_bayes_positive <- trunc_comp_bayes(
  Y ~ R,
  atom = 0,
  data = positive_example,
  continuous_support = "positive_real",
  chains = 4,
  iter_warmup = 500,
  iter_sampling = 1000,
  mixture_components = 6,
  seed = 20260419,
  refresh = 0
)

coef(fit_bayes_positive)
```

The same density-plot helper works for the positive-support model. In that
case the posterior predictive density is built from the fitted Gamma mixture
instead of the real-line Gaussian mixture.

```{r, fig.width=11, fig.height=5.5, out.width="100%"}
positive_non_atom_range <- range(positive_example$Y[positive_example$A == 1])
positive_density_x <- seq(
  0,
  positive_non_atom_range[2] + 1,
  length.out = 300
)

posterior_density_plot(fit_bayes_positive, x = positive_density_x)
```

Posterior predictive checks are also available for the positive-support fit.
The atom check assesses the Bernoulli atom model, while the non-atom check
compares replicated draws from the fitted Gamma mixture with the positive
non-atom outcomes. For the positive-support fit, the non-atom check is
shown on the `log(Y)` scale so the density overlay is not distorted by the
boundary at zero, and the corresponding non-atom posterior predictive p-value
is computed on that same `log(Y)` scale.

```{r}
posterior_predictive_pvalues(fit_bayes_positive)
```

```{r, fig.width=11, fig.height=5.5, out.width="100%"}
ppc_plots_positive <- posterior_predictive_check(fit_bayes_positive, seed = 1)
print(ppc_plots_positive$atom)
print(ppc_plots_positive$continuous)
```

The posterior draws for the positive-support fit can be summarized in the same
way as for the real-line example:

```{r}
posterior_probabilities_positive <- c(
  `P(delta > 0)` = mean(fit_bayes_positive$draws$delta > 0),
  `P(mu_delta > 0)` = mean(fit_bayes_positive$draws$mu_delta > 0),
  `P(delta_atom < 0)` = mean(fit_bayes_positive$draws$delta_atom < 0)
)

posterior_probabilities_positive
```

## Bounded outcomes

Two experimental bounded supports are available. Use
`continuous_support = "bounded_continuous"` for genuinely continuous or finely
measured bounded outcomes with non-atom values strictly inside the supplied
range. Use `continuous_support = "bounded_score"` for integer or grid-valued
clinical scores. Posterior predictive non-atom replications are then sampled on
the same reported score support as the data.

The bounded models keep support and kernel separate. The default
`bounded_kernel = "beta"` uses a bounded Beta mixture. Setting
`bounded_kernel = "logit_normal"` uses independent component locations on the
logit scale with logit-normal kernels. For bounded-score outcomes, the heaping
grids describe possible reporting resolutions. For example,
`heaping_grids = c(1, 5, 10)` means the reported score may arise from exact
integer reporting, rounding to a multiple of five, or rounding to a multiple
of ten. The heaping weights are estimated from the data.

For a data frame named `score_data` with a coded score `Y`, treatment indicator
`R`, and atom `-1`, the call is

```{r bounded-score-example, eval=FALSE}
fit_bayes_score <- trunc_comp_bayes(
  Y ~ R,
  atom = -1,
  data = score_data,
  continuous_support = "bounded_score",
  bounded_kernel = "logit_normal",
  score_min = 0,
  score_max = 100,
  score_step = 1,
  heaping_grids = c(1, 5, 10),
  heaping = "shared",
  chains = 4,
  iter_warmup = 1000,
  iter_sampling = 1000,
  refresh = 0
)

fit_bayes_score$diagnostics[c(
  "divergences", "max_rhat", "min_bulk_ess", "min_tail_ess",
  "core_ok", "truncation_ok", "diagnostic_ok"
)]

posterior_predictive_check(fit_bayes_score, type = "continuous")
```
