---
title: "Likelihood-based analysis of the packaged example data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Likelihood-based analysis of the packaged example data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`TruncComp` provides two likelihood-based frequentist procedures for comparing
two groups when a continuous outcome may be undefined because a substantive
event has occurred:

- `method = "lrt"` for the parametric likelihood ratio analysis
- `method = "splrt"` for the semiparametric likelihood ratio analysis

This vignette applies both methods to the packaged example data set used
throughout the package documentation. The goal is to compare the fitted
probabilities outside the atom, the fitted non-atom mean contrast, and the
coded-endpoint contrast.

## Load the example data

```{r}
library(TruncComp)

data("trunc_comp_example", package = "TruncComp")
example_data <- transform(
  trunc_comp_example,
  A = as.integer(Y != 0)
)

head(example_data, 10)
```

The packaged example uses `0` as the atom. Thus `A = 0` marks the substantive
event for which the continuous outcome is undefined, and `A = 1` marks rows
outside the atom.

It is useful to inspect the example on three scales before fitting either
procedure:

- the atom probability
- the mean among the non-atom outcomes
- the mean of the coded endpoint

```{r}
arm_summary <- do.call(
  rbind,
  lapply(split(example_data, example_data$R), function(d) {
    data.frame(
      n = nrow(d),
      atom_probability = mean(d$A == 0L),
      outside_atom_probability = mean(d$A == 1L),
      mean_non_atom = mean(d$Y[d$A == 1L]),
      mean_coded_endpoint = mean(d$Y),
      row.names = NULL
    )
  })
)

arm_summary$arm <- c("control", "treatment")
arm_summary <- arm_summary[, c("arm", "n", "atom_probability",
                               "outside_atom_probability", "mean_non_atom",
                               "mean_coded_endpoint")]
arm_summary
```

This table makes the target quantities concrete. The two likelihood-based
procedures below use the same atom and the same binary treatment indicator, but
they differ in how strongly they model the non-atom outcome distribution.

## Fit the parametric and semiparametric procedures

The same formula interface fits either the parametric likelihood ratio
procedure or the semiparametric likelihood ratio procedure:

```{r}
fit_lrt <- trunc_comp(
  Y ~ R,
  atom = 0,
  data = example_data,
  method = "lrt"
)

fit_splrt <- trunc_comp(
  Y ~ R,
  atom = 0,
  data = example_data,
  method = "splrt"
)

fit_lrt
fit_splrt
```

The two printed objects report the same three fitted contrasts:

- `mu_delta`: treatment-minus-control non-atom mean contrast
- `alpha_delta`: treatment-to-control odds ratio for being outside the atom
- `delta`: treatment-minus-control coded-endpoint mean contrast

Each fit also stores the joint likelihood ratio statistic and its p-value.

## Compare the fitted likelihood results

```{r}
fit_comparison <- data.frame(
  method = c("Parametric likelihood", "Semiparametric likelihood"),
  mu_delta = c(fit_lrt$mu_delta, fit_splrt$mu_delta),
  alpha_delta = c(fit_lrt$alpha_delta, fit_splrt$alpha_delta),
  delta = c(fit_lrt$delta, fit_splrt$delta),
  statistic = c(fit_lrt$statistic, fit_splrt$statistic),
  p_value = c(fit_lrt$p.value, fit_splrt$p.value),
  row.names = NULL
)

fit_comparison
```

For this example the fitted contrasts are the same to the displayed precision.
This agreement is specific to these data. Both methods target the same
contrasts, but the semiparametric procedure places less structure on the
continuous non-atom distribution.

The full printed summaries also expose the stored component confidence
intervals and the joint likelihood ratio result:

```{r}
summary(fit_lrt)
summary(fit_splrt)
```

Notice that the summary output reports the point estimate for the
coded-endpoint contrast `delta`, but leaves its confidence interval blank.
That is intentional. Unlike `mu_delta` and `alpha_delta`, the package does not
store one default interval for `delta` on the fitted object because there are
multiple supported constructions for that interval, and they answer slightly
different questions. Instead, `delta` intervals are computed on demand through
`confint()` after you choose the interval type explicitly.

## Component confidence intervals

The component intervals are available directly through `confint()`:

```{r}
confint(fit_lrt)
confint(fit_splrt)
```

These correspond to the two primary model components:

- the non-atom mean contrast
- the odds ratio for being outside the atom

Their interval constructions differ by method:

- For `method = "lrt"`, both intervals invert their parametric component
  likelihood ratio statistics. The logistic interval is exponentiated to the
  odds ratio scale.
- For `method = "splrt"`, the interval for `mu_delta` inverts the empirical
  likelihood statistic for the non-atom outcomes, while the interval
  for `alpha_delta` inverts the logistic likelihood ratio statistic and is
  exponentiated to the odds ratio scale.

For this example the fitted component intervals are again very similar across
the two likelihood paths.

## Coded-endpoint intervals for `delta`

Unlike the component intervals, intervals for the coded-endpoint contrast
`delta` are computed on demand. The package provides three choices:

- `method = "welch"` for a direct interval on the coded endpoint
- `method = "profile"` for a profile likelihood interval
- `method = "projected"` for a full likelihood-region projection interval

The likelihood-based `delta` constructions use the full parameter vector of
both arm means and both arm logits. This is necessary because `delta` depends
on the baseline levels as well as the two treatment contrasts. Therefore these
intervals are not obtained from the two-dimensional surface returned by
`joint_contrast_surface()`.

The projected interval optimizes `delta` over the full region defined by a
four-degree-of-freedom likelihood ratio cutoff. The profile interval imposes a
fixed scalar `delta`, profiles every remaining coordinate, and inverts the
result at the one-degree-of-freedom cutoff.

```{r}
extract_delta_interval <- function(fit, method) {
  capture.output(
    interval <- confint(
      fit,
      parameter = "delta",
      method = method,
      plot = FALSE
    )
  )

  data.frame(
    lower = unname(interval[1, 1]),
    upper = unname(interval[1, 2]),
    row.names = NULL
  )
}

delta_interval_table <- do.call(
  rbind,
  lapply(
    list(
      list(label = "Parametric likelihood", fit = fit_lrt),
      list(label = "Semiparametric likelihood", fit = fit_splrt)
    ),
    function(entry) {
      do.call(
        rbind,
        lapply(c("welch", "profile", "projected"), function(method_name) {
          interval <- extract_delta_interval(entry$fit, method_name)
          data.frame(
            method = entry$label,
            interval_type = method_name,
            lower = interval$lower,
            upper = interval$upper,
            row.names = NULL
          )
        })
      )
    }
  )
)

delta_interval_table
```

On this data set the Welch interval is identical across the two fitted objects
because it depends only on the coded endpoint. The profile and projected
intervals use the full four-parameter criterion for the arm means and arm
logits and can differ slightly between the parametric and semiparametric
procedures.

## Joint confidence surfaces

For successful adjusted and unadjusted fits, `TruncComp` also exposes the joint
likelihood ratio surface in `(mu_delta, log_or_delta)` for simultaneous
component inference. In an adjusted fit, the axes are the two treatment
coefficients and the adjustment coefficients are profiled out. The
coded-endpoint intervals above are based on the full four-parameter criterion
rather than this two-dimensional component surface.

Adjusted fits set `delta = NA`, and coded-endpoint intervals are unavailable
because conditional treatment coefficients do not define a standardized
marginal coded-endpoint contrast.

Below is the joint confidence plot for the semiparametric likelihood path on
the same example data:

```{r, fig.width=11, fig.height=5.5, out.width="100%"}
joint_contrast_surface(fit_splrt, resolution = 60)
```
