---
title: "Bayesian analysis of the packaged example data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian analysis of the packaged example data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

cores <- parallel::detectCores(logical = FALSE)
options(mc.cores = if(is.na(cores)) 4L else min(4L, cores))
```

`TruncComp2` now includes an experimental Bayesian interface,
`trunc_comp_bayes()`, for the no-covariate two-group comparison. This vignette
applies that interface to the packaged example data set and shows how to move
from the raw atom-plus-continuous outcome to posterior summaries for the atom
probability contrast, the non-atom mean contrast, and the combined-outcome
contrast. The packaged example uses the default `continuous_support =
"real_line"` pathway, but the same interface also supports
`continuous_support = "positive_real"` when all non-atom outcomes are strictly
positive, and experimental bounded models through
`continuous_support = "bounded_continuous"` or
`continuous_support = "bounded_score"`. The bounded models use Beta mixture
kernels by default and can instead use logit-normal mixture kernels with
`bounded_kernel = "logit_normal"`.

The example data set is intentionally small, but the Stan fit below uses
`4` chains with `500` warmup iterations and `1000` post-warmup iterations per
chain so the posterior summaries are based on a more stable run than a minimal
smoke-test configuration. For an applied analysis, you should still inspect the
sampler diagnostics carefully and increase the sampling effort further if they
remain unsatisfactory.

## Load the example data

```{r}
library(TruncComp2)

data("trunc_comp_example", package = "TruncComp2")
example_data <- transform(
  trunc_comp_example,
  A = as.integer(Y != 0)
)

head(example_data, 10)
```

The packaged example uses `0` as the atom, so `A = 0` marks the undefined or
unobserved outcome and `A = 1` marks rows where the continuous outcome is
observed away from the atom.

Before fitting the Bayesian model, it is helpful to inspect the two groups on
the three scales that matter most:

- the atom probability
- the mean among the non-atom outcomes
- the mean on the original combined-outcome scale

```{r}
arm_summary <- do.call(
  rbind,
  lapply(split(example_data, example_data$R), function(d) {
    data.frame(
      n = nrow(d),
      atom_probability = mean(d$A == 0L),
      observed_probability = mean(d$A == 1L),
      mean_non_atom = mean(d$Y[d$A == 1L]),
      mean_combined = mean(d$Y),
      row.names = NULL
    )
  })
)

arm_summary$arm <- c("control", "treatment")
arm_summary <- arm_summary[, c("arm", "n", "atom_probability",
                               "observed_probability", "mean_non_atom",
                               "mean_combined")]
arm_summary
```

This quick table already shows why the two-part decomposition is useful: one
group can look better on the non-atom mean while also having a worse atom
probability.

## Fit the Bayesian two-part model

The Bayesian interface mirrors the unadjusted formula interface used elsewhere
in the package. In the current implementation:

- the atom is modeled separately in each arm
- the non-atom outcomes are modeled with arm-specific truncated stick-breaking
  mixtures chosen by `continuous_support`
- bounded continuous and bounded score supports can use either Beta kernels or
  logit-normal kernels
- covariate adjustment is not yet supported

```{r fit-bayes, message = FALSE, warning = FALSE}
fit_bayes <- trunc_comp_bayes(
  Y ~ R,
  atom = 0,
  data = example_data,
  chains = 4,
  iter_warmup = 500,
  iter_sampling = 1000,
  mixture_components = 6,
  seed = 20260418,
  refresh = 0
)

fit_bayes
```

The printed fit shows the four main treatment-effect summaries:

- `delta_atom`: difference in atom probability
- `mu_delta`: difference in means among non-atom outcomes
- `alpha_delta`: odds ratio of being observed away from the atom
- `delta`: combined-outcome contrast

The stored posterior summary table is often the most convenient compact view:

```{r}
fit_bayes$summary_table
```

These reported point estimates are posterior medians. The `posterior_prob`
column is the posterior probability that the corresponding contrast exceeds its
null value, which is `0` for `delta_atom`, `mu_delta`, and `delta`, and `1` for
`alpha_delta`.

The model also stores arm-specific posterior summaries:

```{r}
fit_bayes$arm_table
```

Here `rho_0` and `rho_1` are the arm-specific atom probabilities, `pi_0` and
`pi_1` are the arm-specific probabilities of being observed away from the atom,
and `mu_0_c` and `mu_1_c` are the means among the non-atom outcomes.

## Posterior outcome densities

The Bayesian fit object can also be visualized through the arm-specific
posterior outcome densities implied by the fitted DPMM. The plotted curve in
each panel summarizes the posterior predictive density for the observed-outcome
law, so it is weighted by the posterior probability of being away from the
atom. The dark-gray ribbon shows a pointwise credible interval, and the solid
spike with an arrowhead records the atom together with its posterior mean mass.
For this vignette, we supply a compact x-grid directly so the figure focuses on
the region where the example data live while still showing the atom at `0`.

```{r, fig.width=11, fig.height=5.5, out.width="100%"}
observed_range <- range(example_data$Y[example_data$A == 1])
density_x <- seq(
  min(fit_bayes$atom, observed_range[1]) - 0.3,
  observed_range[2] + 0.8,
  length.out = 300
)

posterior_density_plot(fit_bayes, x = density_x)
```

## Posterior predictive checks

The fit object also supports posterior predictive checks for both parts of the
Bayesian model. The atom PPC compares the fitted Bernoulli atom model with the
observed atom indicators, while the continuous PPC compares replicated non-atom
draws from the fitted mixture model with the observed non-atom outcomes. The
package also reports discrepancy-based posterior predictive p-values for these
two checks. These are posterior predictive model-checking summaries rather than
frequentist hypothesis-test p-values.

```{r}
posterior_predictive_pvalues(fit_bayes)
```

```{r, fig.width=11, fig.height=5.5, out.width="100%"}
ppc_plots <- posterior_predictive_check(fit_bayes, seed = 1)
print(ppc_plots$atom)
print(ppc_plots$continuous)
```

## Credible intervals and custom posterior summaries

You can extract equal-tail credible intervals for the main contrasts with
`confint()`:

```{r}
confint(fit_bayes, parameter = c("delta_atom", "mu_delta", "delta"))
```

The full posterior draws are also retained in `fit_bayes$draws`, so it is easy
to compute custom posterior summaries. For example, if smaller atom probability
is preferable, then a treatment benefit on the atom component corresponds to
`delta_atom < 0` rather than `delta_atom > 0`.

```{r}
posterior_probabilities <- c(
  `Pr(delta > 0)` = mean(fit_bayes$draws$delta > 0),
  `Pr(mu_delta > 0)` = mean(fit_bayes$draws$mu_delta > 0),
  `Pr(delta_atom < 0)` = mean(fit_bayes$draws$delta_atom < 0)
)

posterior_probabilities
```

This is one of the main advantages of the Bayesian fit object: the package
returns the posterior draws needed for follow-up summaries rather than only a
single test statistic or p-value.

## Diagnostics and practical tuning

The fit object stores a compact diagnostic summary:

```{r}
data.frame(
  divergences = fit_bayes$diagnostics$divergences,
  max_rhat = fit_bayes$diagnostics$max_rhat,
  min_bulk_ess = fit_bayes$diagnostics$min_bulk_ess,
  min_tail_ess = fit_bayes$diagnostics$min_tail_ess,
  core_ok = fit_bayes$diagnostics$core_ok,
  truncation_ok = fit_bayes$diagnostics$truncation_ok,
  diagnostic_ok = fit_bayes$diagnostics$diagnostic_ok
)
```

For truncated stick-breaking mixtures, `truncation_ok` is based on ordinary
sampler diagnostics for the concentration parameters and the final retained
stick weights.

Because this vignette still uses a small packaged example data set, these
diagnostics should be interpreted as an illustration of the workflow rather
than as definitive evidence that the model is well calibrated for every
application. In a real analysis, a good next step is to rerun the model with
additional sampling effort if needed, then inspect `summary(fit_bayes)` and
`fit_bayes$diagnostics` again.

The same pattern scales directly to user data with the same structure:

1. code the atom explicitly
2. fit `trunc_comp_bayes(Y ~ R, atom = ...)`
3. inspect `summary_table`, `arm_table`, and diagnostics
4. compute any study-specific posterior probabilities from `fit$draws`

## Positive-support continuous outcomes

If the non-atom outcomes are constrained to the positive real line, set
`continuous_support = "positive_real"`. That switches the continuous component
from a Gaussian mixture to a Gamma mixture while keeping the same atom model
and the same posterior estimands.

```{r fit-bayes-positive, message = FALSE, warning = FALSE}
positive_example <- simulate_truncated_data(
  20,
  f0 = function(n) stats::rgamma(n, shape = 4, rate = 2.5),
  f1 = function(n) stats::rgamma(n, shape = 5, rate = 2.2),
  pi0 = 0.7,
  pi1 = 0.6,
  atom = 0
)

fit_bayes_positive <- trunc_comp_bayes(
  Y ~ R,
  atom = 0,
  data = positive_example,
  continuous_support = "positive_real",
  chains = 4,
  iter_warmup = 500,
  iter_sampling = 1000,
  mixture_components = 6,
  seed = 20260419,
  refresh = 0
)

coef(fit_bayes_positive)
```

The same density-plot helper works for the positive-support model. In that
case the posterior predictive density is built from the fitted Gamma mixture
instead of the real-line Gaussian mixture.

```{r, fig.width=11, fig.height=5.5, out.width="100%"}
positive_observed_range <- range(positive_example$Y[positive_example$A == 1])
positive_density_x <- seq(
  0,
  positive_observed_range[2] + 1,
  length.out = 300
)

posterior_density_plot(fit_bayes_positive, x = positive_density_x)
```

Posterior predictive checks are also available for the positive-support fit.
As before, the atom PPC checks the Bernoulli atom model, while the continuous
PPC compares replicated non-atom draws from the fitted Gamma mixture with the
observed positive outcomes. For the positive-support fit, the continuous PPC is
shown on the `log(Y)` scale so the density overlay is not distorted by the
boundary at zero, and the corresponding continuous posterior predictive p-value
is computed on that same `log(Y)` scale.

```{r}
posterior_predictive_pvalues(fit_bayes_positive)
```

```{r, fig.width=11, fig.height=5.5, out.width="100%"}
ppc_plots_positive <- posterior_predictive_check(fit_bayes_positive, seed = 1)
print(ppc_plots_positive$atom)
print(ppc_plots_positive$continuous)
```

The posterior draws for the positive-support fit can be summarized in the same
way as for the real-line example:

```{r}
posterior_probabilities_positive <- c(
  `Pr(delta > 0)` = mean(fit_bayes_positive$draws$delta > 0),
  `Pr(mu_delta > 0)` = mean(fit_bayes_positive$draws$mu_delta > 0),
  `Pr(delta_atom < 0)` = mean(fit_bayes_positive$draws$delta_atom < 0)
)

posterior_probabilities_positive
```

## Bounded Outcomes

Two experimental bounded models are also available. Use
`continuous_support = "bounded_continuous"` for genuinely continuous or finely
measured bounded outcomes with non-atom values strictly inside the supplied
range. Use `continuous_support = "bounded_score"` for integer or grid-valued
clinical scores; posterior predictive non-atom replications are then sampled on
the same reported score support as the observed data.

The bounded models keep support and kernel separate. The default
`bounded_kernel = "beta"` uses the original bounded Beta mixture. Setting
`bounded_kernel = "logit_normal"` uses ordered component locations on the logit
scale with logit-normal kernels. For bounded-score outcomes, the heaping grids
describe possible reporting resolutions. For example, `heaping_grids = c(1, 5,
10)` means the reported score may arise from exact integer reporting, rounding
to a multiple of five, or rounding to a multiple of ten; the heaping weights are
estimated from the data.

```{r bounded-score-example, eval=FALSE}
fit_bayes_score <- trunc_comp_bayes(
  Y ~ R,
  atom = -1,
  data = ist3,
  continuous_support = "bounded_score",
  bounded_kernel = "logit_normal",
  score_min = 0,
  score_max = 100,
  score_step = 1,
  heaping_grids = c(1, 5, 10),
  heaping = "shared",
  mixture_components = 2,
  chains = 4,
  iter_warmup = 300,
  iter_sampling = 300,
  control = list(adapt_delta = 0.99, max_treedepth = 12),
  prior = list(
    rho_alpha = 1,
    rho_beta = 1,
    alpha_shape = 4,
    alpha_rate = 4,
    mu_logit_mean = 0,
    mu_logit_sd = 2,
    sigma_logit_meanlog = log(1.2),
    sigma_logit_sdlog = 0.5,
    eta_prior = c(2, 2, 2)
  ),
  refresh = 0
)

fit_bayes_score$diagnostics[c(
  "divergences", "max_rhat", "min_bulk_ess", "min_tail_ess",
  "core_ok", "truncation_ok", "diagnostic_ok"
)]

posterior_predictive_check(fit_bayes_score, type = "continuous")
```
