---
title: "Likelihood-based analysis of the packaged example data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Likelihood-based analysis of the packaged example data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`TruncComp2` provides two likelihood-based frequentist procedures for the
two-group truncated-outcome problem:

- `method = "lrt"` for the parametric likelihood-ratio analysis
- `method = "splrt"` for the semi-parametric likelihood-ratio analysis

This vignette applies both methods to the packaged example data set used
throughout the package documentation. The goal is to compare the fitted atom
mechanism, the fitted mean contrast among the observed outcomes, and the
combined-outcome contrast on the original outcome scale.

## Load the example data

```{r}
library(TruncComp2)

data("trunc_comp_example", package = "TruncComp2")
example_data <- transform(
  trunc_comp_example,
  A = as.integer(Y != 0)
)

head(example_data, 10)
```

The packaged example uses `0` as the atom, so `A = 0` marks the undefined or
unobserved outcome and `A = 1` marks rows where the continuous outcome is
observed away from the atom.

As in the Bayesian vignette, it is useful to look at the example on the three
most relevant scales before fitting either likelihood:

- the atom probability
- the mean among the non-atom outcomes
- the mean on the original combined-outcome scale

```{r}
arm_summary <- do.call(
  rbind,
  lapply(split(example_data, example_data$R), function(d) {
    data.frame(
      n = nrow(d),
      atom_probability = mean(d$A == 0L),
      observed_probability = mean(d$A == 1L),
      mean_non_atom = mean(d$Y[d$A == 1L]),
      mean_combined = mean(d$Y),
      row.names = NULL
    )
  })
)

arm_summary$arm <- c("control", "treatment")
arm_summary <- arm_summary[, c("arm", "n", "atom_probability",
                               "observed_probability", "mean_non_atom",
                               "mean_combined")]
arm_summary
```

This table makes the target quantities concrete. The two likelihood-based
procedures below use the same atom and the same binary treatment indicator, but
they differ in how strongly they model the non-atom outcome distribution.

## Fit the parametric and semi-parametric likelihoods

For the same formula interface, the package can fit either the parametric
likelihood-ratio procedure or the semi-parametric likelihood-ratio procedure:

```{r}
fit_lrt <- trunc_comp(
  Y ~ R,
  atom = 0,
  data = example_data,
  method = "lrt"
)

fit_splrt <- trunc_comp(
  Y ~ R,
  atom = 0,
  data = example_data,
  method = "splrt"
)

fit_lrt
fit_splrt
```

The two printed objects report the same three fitted treatment-effect summaries:

- `mu_delta`: difference in means among the observed outcomes
- `alpha_delta`: odds ratio of being observed away from the atom
- `delta`: combined-outcome mean difference on the original scale

Each fit also stores the joint likelihood-ratio statistic and its p-value.

## Compare the fitted likelihood results

```{r}
fit_comparison <- data.frame(
  method = c("Parametric likelihood", "Semi-parametric likelihood"),
  mu_delta = c(fit_lrt$mu_delta, fit_splrt$mu_delta),
  alpha_delta = c(fit_lrt$alpha_delta, fit_splrt$alpha_delta),
  delta = c(fit_lrt$delta, fit_splrt$delta),
  statistic = c(fit_lrt$statistic, fit_splrt$statistic),
  p_value = c(fit_lrt$p.value, fit_splrt$p.value),
  row.names = NULL
)

fit_comparison
```

For this particular example the two methods agree extremely closely, and in
fact the fitted contrasts are the same to the displayed precision. That will
not always happen in general. The point of the shared `trunc_comp()` interface
is that both methods target the same treatment-effect summaries even though the
semi-parametric procedure places less structure on the continuous non-atom
distribution.

The full printed summaries also expose the stored component confidence
intervals and the joint likelihood-ratio result:

```{r}
summary(fit_lrt)
summary(fit_splrt)
```

Notice that the summary output reports the point estimate for the
combined-outcome contrast `delta`, but leaves its confidence interval blank.
That is intentional. Unlike `mu_delta` and `alpha_delta`, the package does not
store one default interval for `delta` on the fitted object because there are
multiple supported constructions for that interval, and they answer slightly
different questions. Instead, `delta` intervals are computed on demand through
`confint()` after you choose the interval type explicitly.

## Component confidence intervals

The component intervals are available directly through `confint()`:

```{r}
confint(fit_lrt)
confint(fit_splrt)
```

These correspond to the two primary model components:

- the contrast in mean among the observed outcomes
- the odds ratio for being observed away from the atom

They are not exactly the same type of interval under the two methods.

- For `method = "lrt"`, both component intervals are Wald intervals from the
  fitted parametric submodels: a Normal-model interval for `mu_delta` and a
  logistic-model interval, exponentiated to the odds-ratio scale, for
  `alpha_delta`.
- For `method = "splrt"`, the interval for `mu_delta` is an empirical-
  likelihood interval based on the observed non-atom outcomes, while the
  interval for `alpha_delta` comes from profiling the logistic observation
  model and then exponentiating back to the odds-ratio scale.

For this example the fitted component intervals are again very similar across
the two likelihood paths.

## Combined-outcome intervals for `delta`

Unlike the component intervals, intervals for the combined-outcome contrast
`delta` are computed on demand. The package currently provides three choices:

- `method = "welch"` for a direct interval on the combined outcome
- `method = "profile"` for a profile-likelihood interval
- `method = "projected"` for a joint-surface projection interval

For the two likelihood-based `delta` constructions, `profile` and `projected`,
there are actually two numerical ways to compute the interval:

- a default grid-based route, which evaluates the likelihood surface on a
  finite grid in `(mu_delta, log_or_delta)` space and then profiles or projects
  that surface
- an alternative optimizer-based route, which searches directly for the
  relevant `delta` endpoints without first building the full grid

In the public interface, this is controlled by the `algorithm` argument to
`confint(..., parameter = "delta", method = "profile" | "projected")`. The
default is `algorithm = "grid"`, which is usually the most transparent choice
because it is tied directly to the same joint surface returned by
`joint_contrast_surface()`. The optimizer-based route,
`algorithm = "optimize"`, is available as a slower direct alternative. In this
vignette, the `profile` and `projected` intervals below use that optimizer-
based route.

```{r}
extract_delta_interval <- function(fit, method) {
  algorithm <- if(method %in% c("profile", "projected")) "optimize" else "grid"

  capture.output(
    interval <- confint(
      fit,
      parameter = "delta",
      method = method,
      algorithm = algorithm,
      plot = FALSE
    )
  )

  data.frame(
    lower = unname(interval[1, 1]),
    upper = unname(interval[1, 2]),
    row.names = NULL
  )
}

delta_interval_table <- do.call(
  rbind,
  lapply(
    list(
      list(label = "Parametric likelihood", fit = fit_lrt),
      list(label = "Semi-parametric likelihood", fit = fit_splrt)
    ),
    function(entry) {
      do.call(
        rbind,
        lapply(c("welch", "profile", "projected"), function(method_name) {
          interval <- extract_delta_interval(entry$fit, method_name)
          data.frame(
            method = entry$label,
            interval_type = if(method_name %in% c("profile", "projected")) {
              paste0(method_name, " (optimize)")
            } else {
              method_name
            },
            lower = interval$lower,
            upper = interval$upper,
            row.names = NULL
          )
        })
      )
    }
  )
)

delta_interval_table
```

On this data set the Welch interval is identical across the two fitted objects
because it depends only on the observed combined outcome. The profile and
projected intervals use the fitted likelihood surface and can differ slightly
between the parametric and semi-parametric procedures.

## Joint confidence surfaces

For successful unadjusted fits, `TruncComp2` also exposes the joint
likelihood-ratio surface in `(mu_delta, log_or_delta)`. This is the object that
underlies the projected and profile-style `delta` intervals above.

Below is the joint confidence plot for the semi-parametric likelihood path on
the same example data:

```{r, fig.width=11, fig.height=5.5, out.width="100%"}
joint_contrast_surface(fit_splrt, resolution = 60)
```
